Is structure based drug design ready for selectivity optimization?

### 2018-12-21 manuscript.tex
checking changes in

Alchemical free energy calculations are now widely used to drive or maintain potency in small molecule lead optimization, where the binding affinity to a protein target can be computed---in well-behaved cases---to roughly 1 kcal/mol inaccuracy, which is believed to primarily stem from force field errors.
Despite this, the potential to use free energy calculations to drive optimization of compound \emph{selectivity} among two similar targets has been relatively unexplored.
In the most optimistic scenario, the similarity of binding sites might lead to a fortuitous cancellation of force field errors and allow selectivity to be predicted more accurately than affinity.
Here, we assess the accuracy with which selectivity can be predicted in the context of small molecule kinase inhibitors, considering the very similar binding sites of human kinases CDK2 and CDK9, as well as another series of ligands attempting to achieve selectivity between the more distantly related kinases CDK2 and ERK2.
Using a novel Bayesian analysis approach, we separate force field error from statistical error and quantify the correlation in force field errors between selectivity targets.
We find that, in the closely related CDK2/CDK9 case, a high correlation in force field errors suggests free energy calculations can have significant impact in aiding chemists in achieving selectivity, while in more distantly related kinases (CDK2/ERK2), limited correlation in force field errors reduces the ability for free energy calculations to aid selectivity optimization.
In both cases, the correlation in force field error suggests that longer simulations are beneficial to properly balance statistical error with systematic error to take full advantage of the increase in accuracy in selectivity prediction possible due to fortuitous cancellation of error.

@@

%
%
%  INTRODUCITON (want to see if this works) 
%
%
Free energy methods have proven useful in aiding structure-based drug design by driving the optimization or maintenance of potency in lead optimization. Alchemical free energy calculations allow for prediction of ligand binding free energies, including all enthalpic and entropic contributions~\citep{Chodera2011-jn}. Advances in atomistic molecular mechanics forcefields and free energy methodologies~\citep{Huang:J.Comput.Chem.:2013,Maier:J.Chem.TheoryComput.:2015,Harder:J.Chem.TheoryComput.:2016,Cournia:2017ip} have allowed free energy methods to reach a level of accuracy  sufficient for predicting ligand potencies~\citep{BROWN2009420}. These methods have been applied prospectively to develop inhibitors for Tyk2~\citep{Abel2017-gw}, Syk~\citep{Lovering:2016fg}, BACE1~\citep{Ciordia:2016dn}, GPCRs~\citep{Lenselink:2016ip}, and HIV protease~\citep{Jorgensen:2016dv}. A recent large-scale review found that the use of FEP+~\citep{Wang:J.Am.Chem.Soc.:2015} to predict potency for 92 different projects and 3021 compounds found a median RMSE of 1 kcal/mol~\citep{Abel:2017jt}.

\paragraph{Selectivity is an important consideration in drug design}
In addition to maintaining or optimizing potency, free energy methods can be used to predict the selectivity of a ligand between two or more targets. Selectivity is an important property to consider in drug development, either in the pursuit of a maximally selective inhibitor~\citep{Zhang2009-il,Huggins2012-hr} or in pursuit a polypharmacological agent~\citep{Fan2007-hm,Apsel2008-it,Knight:Nat.Rev.Cancer:2010,Hopkins2006-qu,Hopkins2008-ij}, to avoid on-target toxicity (arising from inhibition of the intended target)~\citep{Rudmann2013-hi}  and off-target toxicity (arising from inhibition of unintended targets)~\citep{Kijima2011-xs,Liu2014-yi}. In either paradigm, considering the selectivity of a compound is complicated by the biology of the target. For example, kinases exist as nodes in complex signaling networks~\citep{Mendoza2011-bj,Tricker2015-xx} with feedback inhibition and cross-talk between pathways. Careful consideration of which off-targets are being inhibited can avoid off-target toxicity due to alleviating feedback inhibition and inadvertently reactivating the targeted pathway~\citep{Mendoza2011-bj,Tricker2015-xx}, or the upregulation of a secondary pathway by alleviation of cross-talk inhibition~\citep{Bailey2014-pd,Chandarlapaty:CancerCell:2011}. Off-target toxicity can also be caused by inhibiting unrelated targets, such as gefitinib, an EGFR inhibitor, inhibiting CYP2D6~\citep{Kijima2011-xs} and causing hepatotoxicity in lung cancer patients. In a cancer setting, on-target toxicity can be avoided by considering the selectivity for the oncogenic mutant form of the kinase over the wild type form of the kinase~\citep{Pao2004-kx,Kim2012-mo,Juchum:DrugResist.Updat.:2015}, exemplified by number of first generation EGFR inhibitors. Selectivity considerations can also lead to beneficial effects: Imatinib, intially developed to target BCR-Abl fusion proteins, is also approved for treating gastrointestinal stromal tumors (GIST)~\citep{Din2008-ag} due to its activity against receptor tyrosine kinase KIT.

\paragraph{Use of physical modeling to predict selectivity is relatively unexplored}
While predicting selectivity is important for drug discovery, the utility of free energy methods for predicting this property is relatively unexplored. If there is fortuitous cancellation of systematic, or forcefield, errors for closely related systems, free energy methods may be much more accurate than expected given the errors made in predicting the potency for each individual target. Molecular dynamics and free energy calculations have been used to extensively study the selectivity of imatinib for Abl kinase over Src~\citep{Lin2013-ft,Lin2014-iv} and within a family of non-receptor tyrosine kinases~\citep{Lin2013-mu}. This work focuses on understanding the role reorganization energy plays in the exquisite selectivity of imatinib for Abl over Src despite high similarity between the cocrystallized binding mode and kinase conformations, and does not touch on the evaluation of the accuracy of these methods, or their application to drug discovery on congeneric series of ligands. Previous work predicting the selectivity of three bromodomain inhibitors accross the bromodomain family achieved promising accuracy for single target potencies of roughly 1 kcal/mol, but does not explicitly evaluate any selectivity metrics~\citep{Aldeghi2017-ox} or look at correlation in the errors made for each bromodomain.

\paragraph{Kinases are an interesting and particularly challenging model system for selectivity predictions}
Kinases are a useful model system to work with for assessing the utility of free energy calculations to predict selectivity. With the approval of imatinib for the treatment of chronic myelogenous leukemia in 2001, targeted small molecule kinase inhibitors (SMKIs) have become a major class of therapeutics in treating cancer and other diseases. 
Currently, there are 43 FDA-approved SMKIs~\citep{fda-approved-kinase-inhibitors}, and it is estimated that kinase targeted therapies account for as much as 50\% of current drug development~\citep{Santos:Nat.Rev.DrugDiscov.:2016}, with many more compounds currently in clinical trials. 
While there have been a number of successes, the current stable of FDA-approved kinase inhibitors targets only a small number of kinases implicated in disease, and the design of new selective kinase inhibitors remains a significant challenge.
Achieving desired selectivity profiles is particularly difficult for kinase targets, making them a system where physical modeling has the potential for a large impact. Achieving selective inhibition of kinases is challenging as, there are more than 518 protein kinases~\citep{Volkamer2015-jx,Manning2002-cw} with a highly conserved ATP binding site that is targeted by the majority of SMKIs~\citep{Wu2015-oq}.
While kinase inhibitors have been designed to target kinase-specific subpockets and binding modes to achieve selectivity~\citep{Cowan-Jacob2007-rn,Seeliger2007-jn,Huse2002-ml,Harrison2003-ct,Volkamer2016-sj,ChristmannFranck:2016gka}, previous work has shown that both Type I (binding to the active, DFG-in conformation) and Type II (binding to the inactive, DFG-out conformation) inhibitors display a range of selectivities~\citep{Anastassiadis2011-sm,Davis:Nat.Biotechnol.:2011}, often exhibiting significant binding to a number of other targets in addition to their primary target. 
Even FDA-approved inhibitors---often the result of extensive drug development programs---bind to a large number of off-target kinases~\citep{Klaeger2017-jr}. Kinases are also targets of interest for developing polypharmacological compounds, or inhibitors that are specifically designed to inhibit multiple kinase targets. Resistance to MEK inhibitors in KRAS-mutant lung and colon cancer has been shown to be driven by HER3 upregulation~\citep{Sun:2014kt}, providing rational for dual MEK/ERBB family inhibitors. Similarly, combined MEK and VEGFR1 inhibition has been proposed as a combinatorial approach to treat KRAS-mutant lung cancer~\citep{Manchado:2016fb}. Developing inhibitors with the desired polypharmacology means navigating more complex selectivity profiles, presenting a problem where physical modeling has the potential to dramatically speed up drug discovery. In well-behaved kinase systems, free energy calculation potency predictions have achieved mean unsigned errors of less than 1.0 kcal/mol~\citep{Wang:J.Am.Chem.Soc.:2015,Abel2017-gw}, suggesting that kinases can be computationally tractable as well as clinically interesting.

\paragraph{Assessing the ability of alchemical free energy methods to predict selectivity}
We anticipate difficulty in predicting selectivity if the errors in the alchemical free energy calculations for two targets are largely uncorrelated, or even anticorrelated. However, correlation in the forcefield errors of the free energies for the two targets could lead to a fortuitous cancellation of errors, making selectivity predictions \emph{more} accurate than potency predictions. Such correlation could occur because the same chemical elements appear in the ligand and in highly related binding sites. 
Here, we investigate the magnitude of this correlation ($\rho$) and the utility of alchemical free energy calculations for the prediction of selectivity. We employed state of the art relative free energy calculations~\citep{Wang:J.Am.Chem.Soc.:2015,Abel:2017jt} to predict the selectivities of two different congeneric ligand series~\citep{Shao2013-oe, Blake2016-su}, as well as present simple numerical models to quantify the potential speed up in selectivity optimization expected for different combinations of per target forcefield errors and correlation coefficient values. To tease out the effects of a limited number of experimental measurements, we develop a new Bayesian approach to quantify the uncertainty in the correlation coefficient in the predicted change in selectivity on ligand modification, incorporating all sources of uncertainty and correlation in the computation to separate statistical from force field error. We find that in the closely related systems of CDK2 and CDK9, a high correlation of force field errors suggests that free energy methods can have a significant impact on speeding up selectivity optimization. Even in the more distantly related case (CDK2/ERK2), correlation in the forcefield errors allows free energy calculations to speed up selectivity optimization, suggesting that these methodologies can impact drug discovery even when comparing systems that are not closely related. We present a model of the impact of per target statistical error at different levels of forcefield error correlation, suggesting that it is worthwhile to expend more effort sampling in systems with high correlation.

%
%
%  RESULTS
%
%
\section{Methods}

\paragraph{Numerical model of selectivity optimization speedup}
To model the impact correlation of forcefield error would have on the expected uncertainty for selectivity predictions in the regime of infinite sampling and zero statistical error, $\sigma_{selectivity}$ was calculated using Equation~\ref{eq11} for 1000 evenly spaced values of the correlation coefficient ($\rho$) from 0 to 1, for a number of combinations of per target forcefield errors ($\sigma_{ff,1}$ and $\sigma_{ff,2}$)

\begin{equation}\label{eq11}
\sigma_{selectivity} = \sqrt{\sigma_{ff,1}^2 + \sigma_{ff,2}^2 - 2\rho\sigma_{ff,1}\sigma_{ff,2}}
\end{equation}

The speed up in selectivity optimization that could be expected from using free energy calculations of a particular per target error ($\sigma_{selectivity}$) was quantified as follows using NumPy (v 1.14.2). An original, true distribution for the changed in selectivity of 200000000 new compounds proposed with respect to a reference compound was modeled as a normal distribution centered around 0 with a standard deviation of 1 kcal/mol. This assumption was made on the basis that the majority of selectivity is driven by the scaffold, and R group modifications will do little to drive changes in selectivity. The 1 kcal/mol distribution is supported by the standard deviations of the selectivity in the experimental datasets referenced in this work, which are all less than, but close, to 1 kcal/mol.

Each of these proposed compounds were "screened" by a free energy calculation technique with a per target forcefield error ($\sigma_{ff}$) of 1 kcal/mol and a specified correlation coefficient $\rho$. A $\sigma_{selectivity}$ was calculated according to Equation~\ref{eq11}. The noise of the computational method was modeled as a normal distribution centered around 0 with a standard deviation of $\sigma_{selectivity}$ and added to the "true" change in selectivity, giving us the predicted change in selectivity ($\Delta S_{compound}$). This process can be described by Equation~\ref{eq14}:

\begin{equation}\label{eq14}
\Delta S_{compound} = \mathcal{N}_{true}(\mu =0, \sigma^2 = 1) + \mathcal{N}_{forcefield}(\mu =0, \sigma_{selectivity}^2(\rho))
\end{equation}

Any compound predicted to have an improvement in selectivity of 1.4 kcal/mol (1 log unit) would then be made and have its selectivity experimental measured, using an experimental method with perfect accuracy. The speedup value for each value of $\rho$ is calculated as the proportion of compounds made with a true selectivity gain of 1.4 kcal/mol divided by the proportion of compounds with a 1.4 kcal/mol improvement in the original distribution, where all of the compounds were made.

This process was repeated for a 100x (2.8 kcal/mol, 2 log unit) selectivity optimization and 50 linearly spaced values of the correlation coefficient ($\rho$) between 0 and 1, for four values of $\sigma_{selectivity}$ and 40000000 compounds in the original distribution.

\paragraph{Numerical model of impact of statistical error on selectivity optimization}
To model the impact of statistical error on selectivity optimization at different levels of correlation in the forcefield error, a similar scheme as above was used. An original, true distribution of 40000000 compounds was proposed with respect to a reference compound, drawing from a normal distribution centered on 0 with a standard deviation of 1 (Numpy v 1.14.2). Each of these proposed compounds were "screened" by a free energy calculation technique with a per target forcefield error ($\sigma_{ff}$) of 0.9 kcal/mol~\citep{Harder2016-zn} and a specified correlation coefficient $\rho$, which was evenly spaced between 0 and 1 in 50 steps.  A $\sigma_{selectivity}$ was calculated according to Equation~\ref{eq11}. Additionally, a per target statistical error ($\sigma_{stat}$) was as in Equation~\ref{eq15}

\begin{equation}\label{eq15}
\sigma_{stat} = \sqrt{\frac{2\sigma^2}{N}}
\end{equation}

Where N is the effort put into running sampling the calculation and $\sigma$ is such that when N is 1, $\sigma_{stat}$ = 0.2 kcal/mol. The statistical error was propagated assuming it was uncorrelated, such as in Equation~\ref{eq11} where $\rho$ = 0, giving us $\sigma_{statistics}$ The forcefield and statistical errors where modeled as Gaussian noise added to the true distribution, as in Equation~\ref{eq16}.

 \begin{equation}\label{eq16}
 \Delta S_{compound} = \mathcal{N}_{true}(\mu =0, \sigma^2 = 1) + \mathcal{N}_{forcefield}(\mu =0, \sigma_{selectivity}^2(\rho)) + \mathcal{N}_{stat}(\mu =0, \sigma_{statistics}^2(\rho))
 \end{equation}

\begin{equation}\label{eq2}
MUE = \frac{ \sum_{0}^{n} \mid \Delta G _{calc} - \Delta G _{exp} \mid}{n}
 \end{equation}

 \begin{equation}\label{eq3}
RMSE = \frac{ \sum_{0}^{n} \sqrt{\Delta~G_{calc}^2 - \Delta~G_{exp}^2}}{n} 
 \end{equation}

 \begin{equation}\label{eq4}
G^{phase}_{i,target} = \mathcal{N}(\mu=0,~sd=25.0~kcal/mol)
 \end{equation}

 For each edge of the FEP map (ligand $i$ --> ligand $j$), there is a contribution from dummy atoms, that was modeled as in Equation~\ref{eq5}.  
  \begin{equation}\label{eq5}
c_{i,j} = \mathcal{N}(\mu=0,~sd=25.0~kcal/mol)
 \end{equation}

 The model was restrained by including data from the FEP+ calculation.

 \begin{equation}\label{eq6}
 \Delta G^{BAR}_{phase, ~ij, ~target} = \mathcal{N}(G^{phase}_{j, target} - G^{phase}_{i, target},~\delta^2\Delta G^{BAR}_{phase, ~ij, ~target}, ~observed = \Delta G^{calc}_{phase,~ij,~target})
 \end{equation}

 Where $~\delta^2\Delta G^{BAR}_{phase, ~ij, ~target}$ is the reported BAR uncertainty from the calculation, and $\Delta G^{calc}_{phase,~ij,~target}$ is the BAR estimate of the free energy for the perturbation between ligands $i$ and $j$ in a given phase.

 From this, we can calculate the $\Delta \Delta G^{FEP}$ for each edge as in Equation~\ref{eq7}:

 \begin{equation}\label{eq7}
 \Delta\Delta G^{FEP}_{target,~ij} = \Delta G^{BAR}_{complex,~ij,~target} - \Delta G^{BAR}_{solvent,~ij,~target}
 \end{equation}

 To model the way an offset is calculated for the $\Delta~G$ reported by the FEP+ panel in Maestro:

\begin{equation}\label{eq8}
\text{offset} =  \frac{\sum^n G^{complex}_{i,target} - G^{solvent}_{i,target}}{n} - \frac{\sum^n \Delta G^{exp}_{i}}{n}
\end{equation}

The offset was added to each $\Delta G^{BAR}_{i}$ to calculate $\Delta G^{sch}_{i}$.

The experimental binding affinity was treated as a true value ($\Delta G^{true}_{i,target}$) corrupted by experimental uncertainty, which is assumed to be 0.3 kcal/mol~\citep{BROWN2009420}, with the values reported in the papers ($\Delta G^{obs}_{i,target}$) treated as observations from this distribution (Equation~\ref{eq12})

\begin{equation}\label{eq12}
\Delta G^{exp}_{i,target} = \mathcal{N}(\Delta G^{true}_{i,target}, 0.3 ~\text{kcal/mol},~observed = \Delta G^{obs}_{i,target})
\end{equation}

$\Delta G^{true}_{i,target}$ was assigned a weak normal prior, as in equation~\ref{eq13}.

\begin{equation}\label{eq13}
\Delta G^{true}_{i,target} = \mathcal{N}(0, ~50~\text{kcal/mol})
\end{equation}

The error for a given ligand was calculated as in Equation~\ref{eq9}.

\begin{equation}\label{eq9}
\epsilon_i = \Delta G^{sch}_i - \Delta G^{true}_i
\end{equation}

From these $\epsilon$ values, we calculated the correlation coefficient, $\rho$ as in Equation~\ref{eq10}. 
\begin{equation}\label{eq10}
\rho = \frac{cov(\epsilon_{target1}, \epsilon_{target2})}{\sigma_{target 1}\sigma_{target 2}}
\end{equation}

 Where $\sigma$ is the standard deviation of $\epsilon$. To quantify $\rho$ for the CDK2/ERK2 calculations, the default NUTS sampler with jitter+adapt\_diag initialization, 1000 tuning steps, and a target accept probability of 0.8 was used to draw 10000 samples from the model. The CDK2/CDK9 model was sampled 20000 times using default NUTS sampler with jitter+adapt\_diag initialization and 3000 tuning steps.

%
%  Results 
%
%

\section{Results}
\paragraph{Free energy methods can be used to predict the selectivity of a compound}
While ligand potency for a single target is often quantified as a free energy of binding ($\Delta G_{binding}$), there are a number of different metrics for quantifying the selectivity of a compound~\citep{Bosc:2017gs,Cheng2010-ip}. Here, we propose a more granular view of selectivity: the change in free energy of binding for a given ligand between two different targets ($\Delta \Delta G_{selectivity}$), which can be calculated as in Equation~\ref{selectivity}. $\Delta \Delta G_{selectivity}$ is a useful measure of compound selectivity once a single, or small panel, of off-targets have been identified.

\begin{equation}\label{selectivity}
\Delta \Delta G_{selectivity} = \Delta G_{\text{binding, target 2}} - \Delta G_{\text{binding, target 1}}
\end{equation}

To predict the $\Delta \Delta G_{selectivity}$ of a compound, we developed a protocol that uses a relative free energy calculation (FEP+)~\citep{Wang:J.Am.Chem.Soc.:2015} to run a map of perturbations between ligands in a congeneric series, as described in depth in the methods section. The calculation is repeated for each target of interest, with identical perturbations (edges) between each ligand (nodes). Each edge represents a relative free energy calculation that quantifies the $\Delta \Delta G$ between the ligands, or nodes. By using provided experimental data, we can convert the $\Delta \Delta G$ from each edge to a single potency value for each value against that target ($\Delta G_{target}$). From this sets of calculations, we can calculate a $\Delta \Delta G_{selectivity}$ for each ligand given two targets of interest. Previous work shows that FEP+ can achieve an accuracy ($\sigma_{target}$) of roughly 1 kcal/mol when predicting potency, which is a combination of systematic forcefield and random statistical error~\citep{Wang:J.Am.Chem.Soc.:2015}. However, it is possible that the forcefield component of that error ($\sigma_{ff}$) may fortuitously cancel when computing $\Delta \Delta G_{selectivity}$, leading to a forcefield component of the selectivity uncertainty that is lower than would be expected.

@@

>>>>

Alchemical free energy calculations are now widely used to drive or maintain potency in small molecule lead optimization, where the binding affinity to a protein target can be computed---in well-behaved cases---to roughly 1 kcal/mol inaccuracy, which is believed to primarily stem from force field errors.
Despite this, the potential to use free energy calculations to drive optimization of compound \emph{selectivity} among two similar targets has been relatively unexplored.
In the most optimistic scenario, the similarity of binding sites might lead to a fortuitous cancellation of force field errors and allow selectivity to be predicted more accurately than affinity.
Here, we assess the accuracy with which selectivity can be predicted in the context of small molecule kinase inhibitors, considering the very similar binding sites of human kinases CDK2 and CDK9, as well as another series of ligands attempting to achieve selectivity between the more distantly related kinases CDK2 and ERK2.
Using a Bayesian analysis approach, we separate force field error from statistical error and quantify the correlation in force field errors between selectivity targets.
We find that, in the closely related CDK2/CDK9 case, a high correlation in force field errors suggests free energy calculations can have significant impact in aiding chemists in achieving selectivity, while in more distantly related kinases (CDK2/ERK2), limited correlation in force field errors reduces the ability for free energy calculations to aid selectivity optimization.
In both cases, the correlation in force field error suggests that longer simulations are beneficial to properly balance statistical error with systematic error to take full advantage of the increase in accuracy in selectivity prediction possible due to fortuitous cancellation of error.

@@

%
%
%  INTRODUCITON (want to see if this works) 
%
%
Free energy methods have proven useful in aiding structure-based drug design by driving the optimization or maintenance of potency in lead optimization. Alchemical free energy calculations allow for prediction of ligand binding free energies, including all enthalpic and entropic contributions~\citep{Chodera2011-jn}. 
Advances in atomistic molecular mechanics forcefields and free energy methodologies~\citep{Huang:J.Comput.Chem.:2013,Maier:J.Chem.TheoryComput.:2015,Harder:J.Chem.TheoryComput.:2016,Cournia:2017ip} have allowed free energy methods to reach a level of accuracy  sufficient for predicting ligand potencies~\citep{BROWN2009420}. 
These methods have been applied prospectively to develop inhibitors for Tyk2~\citep{Abel2017-gw}, Syk~\citep{Lovering:2016fg}, BACE1~\citep{Ciordia:2016dn}, GPCRs~\citep{Lenselink:2016ip}, and HIV protease~\citep{Jorgensen:2016dv}. 
A recent large-scale review found that the use of FEP+~\citep{Wang:J.Am.Chem.Soc.:2015} to predict potency for 92 different projects and 3021 compounds found a median root mean squared error (RMSE) of 1 kcal/mol~\citep{Abel:2017jt}.

\paragraph{Selectivity is an important consideration in drug design}

%In addition to maintaining or optimizing potency, free energy methods can be used to predict the selectivity of a ligand between two or more targets. 
In addition to potency, selectivity is an important property to consider in drug development, either in the pursuit of an inhibitor that is maximally selective~\citep{Zhang2009-il,Huggins2012-hr} or possesses a desired polypharmacology~\citep{Fan2007-hm,Apsel2008-it,Knight:Nat.Rev.Cancer:2010,Hopkins2006-qu,Hopkins2008-ij}.
Controlling selectivity can be useful not only in avoiding off-target toxicity (arising from inhibition of unintended targets)~\citep{Kijima2011-xs,Liu2014-yi}, but also in avoiding on-target toxicity (arising from inhibition of the intended target) by selectively targeting disease mutations~\citep{Rudmann2013-hi}. 
In either paradigm, considering the selectivity of a compound is complicated by the biology of the target. 
For example, kinases exist as nodes in complex signaling networks~\citep{Mendoza2011-bj,Tricker2015-xx} with feedback inhibition and cross-talk between pathways. 
Careful consideration of which off-targets are being inhibited can avoid off-target toxicity due to alleviating feedback inhibition and inadvertently reactivating the targeted pathway~\citep{Mendoza2011-bj,Tricker2015-xx}, or the upregulation of a secondary pathway by alleviation of cross-talk inhibition~\citep{Bailey2014-pd,Chandarlapaty:CancerCell:2011}. 
Off-target toxicity can also be caused by inhibiting unrelated targets, such as gefitinib, an EGFR inhibitor, inhibiting CYP2D6~\citep{Kijima2011-xs} and causing hepatotoxicity in lung cancer patients. In a cancer setting, on-target toxicity can be avoided by considering the selectivity for the oncogenic mutant form of the kinase over the wild type form of the kinase~\citep{Pao2004-kx,Kim2012-mo,Juchum:DrugResist.Updat.:2015}, exemplified by number of first generation EGFR inhibitors. 
Selectivity considerations can also lead to beneficial effects: Imatinib, intially developed to target BCR-Abl fusion proteins, is also approved for treating gastrointestinal stromal tumors (GIST)~\citep{Din2008-ag} due to its activity against receptor tyrosine kinase KIT.

\paragraph{Use of physical modeling to predict selectivity is relatively unexplored}
While engineering compound selectivity is important for drug discovery, the utility of free energy methods for predicting this property with the aim of reducing the number of compounds that must be synthesized to achieve the desired selectivity has been relatively unexplored. 
If there is fortuitous cancellation of systematic (forcefield) errors for closely related systems, free energy methods may be much more accurate than expected given the errors made in predicting the potency for each individual target. Molecular dynamics and free energy calculations have been used to extensively study the selectivity of imatinib for Abl kinase over Src~\citep{Lin2013-ft,Lin2014-iv} and within a family of non-receptor tyrosine kinases~\citep{Lin2013-mu}. This work focuses on understanding the role reorganization energy plays in the exquisite selectivity of imatinib for Abl over Src despite high similarity between the cocrystallized binding mode and kinase conformations, and does not touch on the evaluation of the accuracy of these methods, or their application to drug discovery on congeneric series of ligands. Previous work predicting the selectivity of three bromodomain inhibitors accross the bromodomain family achieved promising accuracy for single target potencies of roughly 1 kcal/mol, but does not explicitly evaluate any selectivity metrics~\citep{Aldeghi2017-ox} or look at correlation in the errors made for each bromodomain.

\paragraph{Kinases are an interesting and particularly challenging model system for selectivity predictions}
Kinases are a useful model system to work with for assessing the utility of free energy calculations to predict inhibitor selectivity in a drug discovery context. 
With the approval of imatinib for the treatment of chronic myelogenous leukemia in 2001, targeted small molecule kinase inhibitors (SMKIs) have become a major class of therapeutics in treating cancer and other diseases. 
Currently, there are 43 FDA-approved SMKIs~\citep{fda-approved-kinase-inhibitors}, and it is estimated that kinase targeted therapies account for as much as 50\% of current drug development~\citep{Santos:Nat.Rev.DrugDiscov.:2016}, with many more compounds currently in clinical trials. 
While there have been a number of successes, the current stable of FDA-approved kinase inhibitors targets only a small number of kinases implicated in disease, and the design of new selective kinase inhibitors remains a significant challenge.

%Achieving desired selectivity profiles is particularly difficult for kinase targets, making them a system where physical modeling has the potential for a large impact. 
Achieving selective inhibition of kinases is quite challenging, as there are more than 518 protein kinases~\citep{Volkamer2015-jx,Manning2002-cw} sharing a highly conserved ATP binding site that is targeted by the majority of SMKIs~\citep{Wu2015-oq}.
While kinase inhibitors have been designed to target kinase-specific subpockets and binding modes to achieve selectivity~\citep{Cowan-Jacob2007-rn,Seeliger2007-jn,Huse2002-ml,Harrison2003-ct,Volkamer2016-sj,ChristmannFranck:2016gka}, previous work has shown that both Type I (binding to the active, DFG-in conformation) and Type II (binding to the inactive, DFG-out conformation) inhibitors display a range of selectivities~\citep{Anastassiadis2011-sm,Davis:Nat.Biotechnol.:2011}, often exhibiting significant binding to a number of other targets in addition to their primary target. 
Even FDA-approved inhibitors---often the result of extensive drug development programs---bind to a large number of off-target kinases~\citep{Klaeger2017-jr}.
Kinases are also targets of interest for developing polypharmacological compounds, or inhibitors that are specifically designed to inhibit multiple kinase targets. Resistance to MEK inhibitors in KRAS-mutant lung and colon cancer has been shown to be driven by HER3 upregulation~\citep{Sun:2014kt}, providing a rationale for dual MEK/ERBB family inhibitors. Similarly, combined MEK and VEGFR1 inhibition has been proposed as a combinatorial approach to treat KRAS-mutant lung cancer~\citep{Manchado:2016fb}. Developing inhibitors with the desired polypharmacology means navigating more complex selectivity profiles, presenting a problem where physical modeling has the potential to dramatically speed up drug discovery.

\paragraph{Assessing the ability of alchemical free energy methods to predict selectivity}
Since the prediction of selectivity depends on predicting affinities to two or more targets (or relative affinities between pairs of related molecules), a spectrum of possibilities exists for the accuracy of predicted selectivity metrics.
In well-behaved kinase systems, for example, free energy calculation potency predictions have achieved mean unsigned errors of less than 1.0 kcal/mol~\citep{Wang:J.Am.Chem.Soc.:2015,Abel2017-gw}, believed to arise predominantly from systematic force field errors~\cite{schrodinger-1kcalmol-forcefield-error-paper}.
In the best-case scenario, the force field errors for two protein targets might exactly cancel out, allowing selectivity to be predicted more accurately than potency;
on the other hand, if the force field error acts like a random error between two distinct (but potentially related) protein targets, predictions of selectivity will be \emph{less accurate} than potency predictions.
Real-world systems are likely to be somewhere between these extremes, and quantifying the \emph{degree} to which error in multiple protein targets is correlated, its implications for the use of free energy calculations for prioritizing synthesis in the pursuit of selectivity, the ramifications for optimal calculation protocols, and rough guidelines governing which systems we might expect good selectivity prediction is the primary focus of this work.

Here, we investigate the magnitude of the correlation ($\rho$) in predicted binding free energy differences between compounds to two different targets ($\Delta \Delta G$), assessing the utility of alchemical free energy calculations for the prediction of selectivity. 
We employ state of the art relative free energy calculations~\citep{Wang:J.Am.Chem.Soc.:2015,Abel:2017jt} to predict the selectivities of two different congeneric ligand series~\citep{Shao2013-oe, Blake2016-su}, as well as present simple numerical models to quantify the potential speed up in selectivity optimization expected for different combinations of per target forcefield errors and correlation coefficient values. 
To tease out the effects of a limited number of experimental measurements, we develop a new Bayesian approach to quantify the uncertainty in the correlation coefficient in the predicted change in selectivity on ligand modification, incorporating all sources of uncertainty and correlation in the computation to separate statistical from force field error. 
We find that in the closely related systems of CDK2 and CDK9, a high correlation of force field errors suggests that free energy methods can have a significant impact on speeding up selectivity optimization. 
Even in the more distantly related case (CDK2/ERK2), correlation in the forcefield errors allows free energy calculations to speed up selectivity optimization, suggesting that these methodologies can impact drug discovery even when comparing systems that are not closely related. 
We present a model of the impact of per target statistical error at different levels of forcefield error correlation, suggesting that it is worthwhile to expend more effort sampling in systems with high correlation.

%
%
%  RESULTS
%
%

\section{Results}

\paragraph{Free energy methods can be used to predict the selectivity of a compound}

While ligand potency for a single target is often quantified as a free energy of binding ($\Delta G_\text{binding}$), there are a number of different metrics for quantifying the selectivity of a compound~\citep{Bosc:2017gs,Cheng2010-ip}. Here, we propose a more granular view of selectivity: the change in free energy of binding for a given ligand between two different targets ($\Delta \Delta G_\text{selectivity}$), which can be calculated as in Equation~\ref{selectivity}. 
$\Delta \Delta G_\text{selectivity}$ is a useful measure of compound selectivity once a single, or small panel, of off-targets have been identified. 
\begin{equation}\label{selectivity}
\Delta \Delta G_\text{selectivity} = \Delta G_{\text{binding, target 2}} - \Delta G_{\text{binding, target 1}}
\end{equation}
To predict the $\Delta \Delta G_\text{selectivity}$ of a compound, we developed a protocol that uses a relative free energy calculation (FEP+)~\citep{Wang:J.Am.Chem.Soc.:2015} to run a map of perturbations between ligands in a congeneric series, as described in depth in the methods section. The calculation is repeated for each target of interest, with identical perturbations (edges) between each ligand (nodes). Each edge represents a relative free energy calculation that quantifies the $\Delta \Delta G$ between the ligands, or nodes. By using provided experimental data, we can convert the $\Delta \Delta G$ from each edge to a single potency value for each value against that target ($\Delta G_\text{target}$). From this sets of calculations, we can calculate a $\Delta \Delta G_{selectivity}$ for each ligand given two targets of interest. Previous work shows that FEP+ can achieve an accuracy ($\sigma_\text{target}$) of roughly 1 kcal/mol when predicting potency, which is a combination of systematic forcefield and random statistical error~\citep{Wang:J.Am.Chem.Soc.:2015}. However, it is possible that the forcefield component of that error ($\sigma_{ff}$) may fortuitously cancel when computing $\Delta \Delta G_{selectivity}$, leading to a forcefield component of the selectivity uncertainty that is lower than would be expected.

@@

%
%
% METHODS
%
%

\section{Methods}

\paragraph{Numerical model of selectivity optimization speedup}
To model the impact correlation of forcefield error would have on the expected uncertainty for selectivity predictions in the regime of infinite sampling and zero statistical error, $\sigma_\text{selectivity}$ was calculated using Equation~\ref{eq11} for 1000 evenly spaced values of the correlation coefficient ($\rho$) from 0 to 1, for a number of combinations of per target forcefield errors ($\sigma_\text{ff,1}$ and $\sigma_\text{ff,2}$) 
\begin{equation}
\sigma_\text{selectivity} = \sqrt{\sigma_\text{ff,1}^2 + \sigma_\text{ff,2}^2 - 2 \rho \, \sigma_\text{ff,1} \, \sigma_\text{ff,2}} \label{eq11}
\end{equation}
The speed up in selectivity optimization that could be expected from using free energy calculations of a particular per target error ($\sigma_\text{selectivity}$) was quantified as follows using NumPy (v 1.14.2). 
An original, true distribution for the change in selectivity of 200 000 000 new compounds proposed with respect to a reference compound was modeled as a normal distribution centered around 0 with a standard deviation of 1 kcal/mol. 
This assumption was made on the basis that the majority of selectivity is driven by the scaffold, and R group modifications will do little to drive changes in selectivity. The 1 kcal/mol distribution is supported by the standard deviations of the selectivity in the experimental datasets referenced in this work, which are all less than, but close, to 1 kcal/mol.
